---
title: "Predicting Remaining Useful Life of NASA Aircraft Engines"
subtitle: "Using Cox Proportional Hazards Model"
author: "Jayme Reed & Brad Paton (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    theme: spacelab
    toc: true
    toc-depth: 3
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"}

Literature Comments: [Literature.html](articleSummaries.html)

## Introduction

As the COVID-19 pandemic demonstrated, the ability to predict the
survival of a patient based on their symptoms and medical history is a
major benefit as society deals with health crises such as pandemics.
Those predictions would assist medical practitioners in triage and allow
researchers the ability to generate predictions on the individuals who
are at the greatest risk of losing their lives due to an illness. While
the medical world is the most obvious use of survival predictions, the
same methods can be applied to the predicting bank failure, the
remaining useful life of machines, machinery maintenance timelines, and
insurance likelihood of payouts. There are various survival methods that
can be examined, including Cox proportional hazards model which will be
examined in this paper.

Cox proportional hazards (CPH) models are statistical regression models
specializing in modeling time-to-event predictions with survival data.
Survival data is data with a value for time and an event
[@Abeysekera2009]. For example, in a study investigating the survival
likelihood of mechanical equipment, the time value would begin when the
equipment begins functioning and would conclude in the event the
equipment stops functioning [@Smith2003]. CPH models can use single or
multiple covariates as a means of creating predictions until failure and
are capable of dealing with survival data that contains censored data as
well [@Seung2023]. Censored data is when the information about an
individual in a study is only known for a certain period of time, such
as the length of the study [@klein2005, 1]. The models also have the
advantage of being semi-parametric which allows more flexibility as
compared to other models [@Ming-Chiang2014].

As mentioned, CPH is used in various fields, primarily the health
industry, but the usage of the model does not come without challenges. A
COVID-19 study done using data from a Pakistani hospital noted that the
model was limited due to the recovery possibility of COVID-19 patients
as CPH assumes as time goes on, the survival probability will approach
zero with no survivors. Though CPH was not the best model for this
study, the results of the model were that individuals who were
asymptomatic and young had a higher chance of surviving the virus
[@asghar2024]. In 2024, a study was done in the US on the relationship
between chest pain occurring and mortality of the person experiencing
chest pain using CPH. A response to that study was written that provided
the limitations of CPH with the biggest concern being the assumption of
constant hazards ratio that CPH relies on for the model. In the heart
study, the effect of Triglyceride-Glucose was assumed to be a constant,
but is actually a value that fluctuates over time along with an
individual's body weight and blood glucose levels. As such, the model
would struggle to predict correctly the effect of Triglyceride-Glucose
given the violation of the constant hazards ratio [@jiang2024].

An additional limitation of CPH comes from the data in the covariates
selected and how can they can become biased. Two studies done using UK
health data demonstrated how bias can impact the values for the
covariates when relying on patient submitted information. The first was
a study on the connection between salt intake and the development of
anxiety or depression after 14.5 years [@wang2025] and the second was a
study on the association between postpartum depression and the
development of two or more chronic diseases [@zhang2025]. Both of those
studies relied on patient submitted health data and survey responses
that were supplemented with death records and other accessible health
records thus creating a possible bias in their data which could
negatively influence their model as their covariates may not have
accurately reflected the true data. These articles provide examples of
the care that needs to be taken when working with CPH models.

While health care is the dominate field in which CPH is used, a study in
2021 provides an example of using CPH in insurance fields. An insurance
company in the Czech Republic provides policies for certain illness
events until the event occurs when the insured will receive money and
their contract is complete. To assist the company with predicting if an
individual will have one of the illness events occur, a model using CPH
was generated using gender, age, and region as the
covariates[@zapletal2021]. A study in 2023 focused on the length of
productive life of female Floridian goats to determine the factors that
would impact the longevity of the goats. CPH was used to determine that
the age of first kidding and the specific herd impacts the length of
productive life of the goats [@ziadi2023]. Both of these studies provide
examples of the reaches of CPH models outside the health industry.

Continuing with using CPH outside of health data, we will study the
remaining useful life of a NASA aircraft engine using data provided by
NASA for the study on propagation modeling [@saxena2008]. From the
literature review, we have noted that we will need to be concerned with
the constant hazards ratio within our data and determine how we will
accommodate censored data when creating our model.

From NASA's Open Data Portal CMAPSS Jet Engine Simulated Data we will be using the test_FD002 and train_FD002 datasets available for download at
[NASA's Open Data Portal](https://data.nasa.gov/Aerospace/CMAPSS-Jet-Engine-Simulated-Data/ff5v-kuh6/about_data) and also available on the home page of our [GitHub repository](https://github.com/JR-87/capstone4ds_reed).

These datasets contain time series data for the survival time of different engines under different conditions. The conditions are represented by three operational setting fields and 21 senor measurement fields.


Headers of the train and test data are below.

```{r}
testurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/test_FD002.txt'

test <- read.table(testurl, header = TRUE, sep = '')
head(test)
```

```{r}
trainurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/train_FD002.txt'

train <- read.table(trainurl, header = TRUE, sep = '')
head(test)
```

The visual below shows a histogram with distributions of survival times of the 260 unique engines in the training data.


```{r, echo=FALSE}
library(tidyverse)

distributions <- train %>% 
  group_by(id) %>%
  summarise(time = max(time))
  

ggplot(data = distributions,
       aes(x = time)) +
  geom_histogram(binwidth = 5,
                 fill = "grey")+
  labs(title = "Survival Time Distribution",
       x = "Survival Time",
       y = "Frequency")+
  theme_bw()+
  theme(plot.title = element_text(hjust = .5))

```










## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
